---
title: "The lidR package - LAS formal class"
author: "Eduardo González"
date: "23. February 2019"
output:
  html_document:
    df_print: paged
    theme: flatly
---

This materials have been adapted from the official [lidR wiki documentation](https://github.com/Jean-Romain/lidR/wiki) and [A Brief Introduction of lidR](http://xzsunbest.tk/2018/07/30/ABriefIntroductionOfLidR/).


# The `rlas` package

The `rlas` package relies on a modified version of `LASlib` and `LASzip` libraries (by Martin Isenburg) that were modified to be compatible with `R`. See the official documentation of [the `rlas` package](https://cran.r-project.org/web/packages/rlas/index.html) for more information.

The main use of the `rlas` package is to read and write `.las` and `.laz` binary files used to store LiDAR data. LAS version 1.0 to 1.4 are supported. Point data record format 0,1,2,3,6,7,8 are supported.

The tools offer some basic functionality to directly read and manipulate LAS data.

```{r}
library(rlas)
file_name <- "/appl/data/geo/mml/laserkeilaus/2008_latest/2019/L331/1/L3313F3.laz"
lasdata   <- read.las(file_name)
lasheader <- read.lasheader(file_name)
```


## Basic structure of an rlas data object

You can verify that the an `rlas` object is simply a data table with the data for each lidar point as a row:
```{r }
class(lasdata)
names(lasdata)
```

See a subset of the lidar data table:
```{r}
print(lasdata[c(1:10),])
```

The original laz file size is 190Mb, but its size when loaded in memory to R is 3.3Gb. To see the size of the data table in memory:
```{r}
size <- object.size(lasdata)
print(size, units = "auto")
```

You may apply filters to columns and what points to load based on their attributes when loading LAS files:
```{r}
filtered_las <- read.las(file_name, select = "ia", filter = "-keep_first -drop_intensity_below 95")
size <- object.size(filtered_las)
print(size, units = "auto")
```

The example above loads the coordinate columns x, y and z which are always loaded plus the intensity (i) column and the san angle (a) column. Then only the rows representing a first return point are loaded. Check the documentation for this function with `?read.las` or from the [rlas reference manual](https://cran.r-project.org/web/packages/rlas/rlas.pdf) to see all the available values for selection and filtering. Note that the filter values are the same as those in LAStools and can be checked with:
```{r}
rlas:::lasfilterusage()
```

To write the fildered LAS data as a las or compressed laz file you ned to create the header first and then use `write.las()` function A las or laz file is created following the extension you inditate in the command:
```{r}
filtered_header <- header_create(filtered_las)
write.las("./outputs/out_las.las", filtered_header, filtered_las)
write.las("./outputs/out_las.laz", filtered_header, filtered_las)
```

It is recommendable to also create an index for LAS file, you can do that with the `writelax()` function.
```{r}
writelax("./outputs/out_las.laz")
```
